---
title: "Finn Prediction Assignment Writeup"
author: "Chapman Finn"
date: "8/12/2017"
output:
  html_document:
    fig_height: 9
    fig_width: 9
---
## Introduction  
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here:

http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Class Specifications:
Class A : Dumbbell Biceps Curl in five different fashions
Class B : throwing the elbows to the front
Class C : lifting the dumbbell only halfway
Class D : lowering the dumbbell only halfway
Class E : throwing the hips to the front 

In this project, the dataset is collected from accelerometers on the belt, forearm, arm, and dumbell.  


## Data Preparation  
```{r echo=TRUE}
library(randomForest)
library(corrplot)
library(caret)
library(e1071)
library(lattice)
library(ggplot2)
library(rpart)
library(rpart.plot)
```


### Download the Data
```{r echo=TRUE}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFilePath <- "./data/pml-training.csv"
testFilePath  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFilePath)) {
  download.file(trainUrl, destfile=trainFilePath, method="curl")
}
if (!file.exists(testFilePath)) {
  download.file(testUrl, destfile=testFilePath, method="curl")
}
```


### Read the Data
After downloading the data, we can read its content into two data frames.  

```{r echo=TRUE}
trainRaw <- read.csv("./data/pml-training.csv")
testRaw <- read.csv("./data/pml-testing.csv")
dim(trainRaw)
```

```{r echo=TRUE}
dim(testRaw)
```

The training data set contains 19622 observations and 160 variables, while the testing data set contains 20 observations and 160 variables. The "classe" variable in the training set is the outcome to predict. 

The names of all the variables, "classe" being the 160th variable, are called below.

```{r echo=TRUE}
names(trainRaw)
```


### Clean the data
We will keep the relevant variables and get rid of the rest. First, we remove columns that contain NA, or missing, values.

```{r echo=TRUE}
trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0] 
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0] 
```

Next, we get rid of attributes that do not contribute much to the measurements like ^timestamp, which is to purpose warehousing needs and is not relevant for this analysis.

```{r echo=TRUE}
final_class <- trainRaw$classe
trainFilter <- grepl("^X|timestamp|window|username", names(trainRaw))
train_OnlyImpFeatures <- trainRaw[, !trainFilter]
trainCleaned <- train_OnlyImpFeatures[, sapply(train_OnlyImpFeatures, is.numeric)]

trainCleaned$classe <- final_class

testFilter <- grepl("^X|timestamp|window|username", names(testRaw))
test_OnlyImpFeatures <- testRaw[, !testFilter]
testCleaned <- test_OnlyImpFeatures[, sapply(test_OnlyImpFeatures, is.numeric)]

names(trainCleaned)
```

The number of variables has been reduced from 160 to 53. Now, the cleaned training data set contains 19622 observations and 53 variables, while the testing data set contains 20 observations and 53 variables. The "classe" variable still exists in the cleaned training set.


### Data Slicing
Now, we can split the cleaned training set into a pure training data set (70%) and a validation data set (30%). We will use the validation data set to conduct **cross validation** in future steps.

```{r  echo=TRUE}
set.seed(12345) 
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]
```


## Fitting the model
We can now fit a predictive model for activity recognition using the **Random Forest** algorithm because it automatically selects important variables and is robust to correlated covariates & outliers in general. We will set training control parameter to 5 for **5-fold cross validation** when applying the algorithm.  

```{r  echo=TRUE}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```


## Performance Evaluation
Next, we will estimate the performance of the model on the validation data set.  

```{r  echo=TRUE}
predictRf <- predict(modelRf, testData)
confusionMatrix(testData$classe, predictRf)
```

```{r echo=TRUE}
accuracy <- postResample(predictRf, testData$classe)
accuracy
```

Wow! The accuracy is 0.9889550. The out of sample error is 0.011045.


## Predicting on the test Data Set
Now, we apply the model to the original testing data set downloaded from the data source. 

```{r echo=TRUE}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```


## Appendix: Visulaization
1. Decision Tree Visualization

```{r echo=TRUE}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) 
```
